---
output: html_document
editor_options: 
  chunk_output_type: inline
---
# Chapter 6 Analysis of longitudinal data

We use the following R packages: 

```{r, results="hide", message=FALSE, warning=FALSE, error=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

## 6.1. Implement the analyses of Chapter 8 of MABS using the RATS data

Prepare the dataset

```{r}
# read the RATS data
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
# Factor variables ID and Group in RATS data
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
# Convert to long form and add a Time variable to RATS
RATSL <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) %>%
  mutate(Time = as.integer(substr(WD,3,4))) 
```

```{r}
# Check the dimensions of the data
dim(RATSL)

# Plot the RATSL data
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")
```

### Figure 8.1 Individual response profiles by treatment group for the RATS data

```{r}
p1 <- ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
p6
```
### Figure 8.2 Individual response profiles for RATS data after standardization

```{r}
# Standardize the scores:
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate( stdrats = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
glimpse(RATSL)
p1 <- ggplot(RATSL, aes(x = Time, y = stdrats, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(name = "standardized Weight")
p6
```
### Figure 8.3 Mean response profiles for the three treatment groups in the RATS data

```{r}
# Number of subjects (per group):
n <- 20
# Make a summary data:
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean=mean(Weight), se=sd(Weight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATSS)
p1 <- ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group))
p2 <- p1 + geom_line() + scale_linetype_manual(values = c(1,2,3))
p3 <- p2 + geom_point(size=3) + scale_shape_manual(values = c(1,2,3))
p4 <- p3 + geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3)
p5 <- p4 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6 <- p5 + theme(legend.position = c(0.8,0.5))
p7 <- p6 + scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
p7
```
### Figure 8.4 Boxplots for the RATS data

```{r}
p1 <- ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group))
p2 <- p1 + geom_boxplot(position = position_dodge(width = 0.9))
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + theme(legend.position = c(0.8,0.4))
p5 <- p4 + scale_x_discrete(name = "Time")
# Black & White version:
#p6 <- p5 + scale_fill_grey(start = 0.5, end = 1)
p5
```

### Figure 8.5 Boxplots of mean summary measures for the three treatment groups in the RATS data

```{r}
# Make a summary data of the post treatment Times (1-8)
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
glimpse(RATSL8S)
p1 <- ggplot(RATSL8S, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Times 1-8")
p5
```

### Figure 8.6 Boxplots of mean summary measures for the three treatment groups in the RATS data, without the outlier shown in Figure 8.5

```{r}
# Remove the outlier:
RATSL8S1 <- RATSL8S %>%
  filter(mean < 550)
glimpse(RATSL8S1)
p1 <- ggplot(RATSL8S1, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Times 1-8")
p5
```

### Table 8.3 Independent Samples t-test on the Mean Summary Measure for the RATS Data, Without the Outlier Shown in Figure 8.5

```{r}
# Without the outlier, apply Student's t-test, two-sided:
#ERROR
#t.test(mean ~ Group, data = RATSL8S1, var.equal = TRUE)
```

### Table 8.4 Analysis of Covariance of the RATS Data with Baseline RATS and Treatment Groups as Covariates

```{r}
# Add the baseline from the original data as a new variable to the summary data:
baseline <- RATS$WD1
RATSL8S2 <- RATSL8S %>%
  mutate(baseline)
# Fit the ANCOVA model and see the results:
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)
summary(fit)
anova(fit)
```


## 6.2. Implement the analyses of Chapter 9 of MABS using the BPRS data.

Prepare the dataset

```{r}
# Read the BPRS data
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)
# Factor variables 'treatment' and 'subject' in BPRS data
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
# Convert the data sets to long form and add a week variable to BPRS
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject) 
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))
```


### Table 9.1: Bprs Recorded Over a 9-Time Period

```{r, echo=TRUE}
BPRS <- within(BPRS, {
       subject <- factor(subject)
    treatment <- factor(treatment)
})
glimpse(BPRS)
```
### Table 9.2 Long Form of the Data

```{r, echo=TRUE}
glimpse(BPRSL)
head(BPRSL); tail(BPRSL)
```
### Figure 9.1 Plot of bprs against week for bprs data, ignoring the repeated-measures structure of the data but identifying the group to which each observation belongs

```{r, echo=TRUE}
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = subject))
p2 <- p1 + geom_text(aes(label = treatment))
p3 <- p2 + scale_x_continuous(name = "Week", breaks = seq(0, 60, 10))
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw()
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```

### Table 9.3 Fitting a Linear Regression Model to BPRS Data with bprs as Response Variable, and treatment and week as Explanatory Variables, and Ignoring the Repeated-Measures Structure of the Data

```{r, echo=TRUE}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)
```
### Figure 9.2 Plot of individual bprs profiles

```{r, echo=TRUE}
#ERROR
#p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = subject))
#p2 <- p1 + geom_line(aes(linetype = treatment))
#p3 <- p2 + scale_x_continuous(name = "week", breaks = seq(0, 60, 10))
#p4 <- p3 + scale_y_continuous(name = "bprs")
#p5 <- p4 + theme_bw() + theme(legend.position = "top")
#p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
#p6

#Error: geom_path: If you are using dotted or dashed lines, colour, size and linetype must be constant over the line
#Run `rlang::last_error()` to see where the error occurred.
```
```{r}
# Draw the plot
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

### Figure 9.3 Scatterplot matrix of repeated measures in bprs data

```{r, echo=TRUE, fig.width=10, fig.height=10}
pairs(BPRS[, 3:13], cex = 0.7)
```
### Table 9.4 Fitting Random Intercept Model, with week and treatment as Explanatory Variables, to BPRS Data

```{r, echo=TRUE}
library("lme4")
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)

```
### Table 9.5 Fitting the Random Intercept and Slope Model, with week and treatment as Explanatory Variables, to BPRS Data

```{r, echo=TRUE}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
anova(BPRS_ref1, BPRS_ref)
```

### Table 9.6 Fitting the Random Intercept and Slope Model that Allows for a treatment × week Interaction to BPRS Data

```{r, echo=TRUE}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
anova(BPRS_ref1, BPRS_ref2)
```

### Figure 9.4 Fitted bprs profiles from the interaction model and observed bprs profiles

```{r, echo=TRUE, fig.width=3, fig.height=4}
Fitted <- fitted(BPRS_ref2)
BPRSL <- BPRSL %>% mutate(Fitted)
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = subject))
p2 <- p1 + geom_line(aes(linetype = treatment))
p3 <- p2 + scale_x_continuous(name = "week", breaks = seq(0, 60, 20))
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Observed")
graph1 <- p7
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = subject))
p2 <- p1 + geom_line(aes(linetype = treatment))
p3 <- p2 + scale_x_continuous(name = "week", breaks = seq(0, 60, 20))
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Fitted")
graph2 <- p7
#graph1; graph2

#Error: geom_path: If you are using dotted or dashed lines, colour, size and linetype must be constant over the line
#Run `rlang::last_error()` to see where the error occurred.
```





```{r}
# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0).
BPRSL8S <- BPRSL %>%
  filter(week > 0) %>%
  group_by(treatment, subject) %>%
  summarise( mean=mean(bprs) ) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSL8S)

# Draw a boxplot of the mean versus treatment
ggplot(BPRSL8S, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")

# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
BPRSL8S1 <- BPRSL8S %>%
  filter(mean < 60)
```





_This is the end of the report. That's all folks!_

```{r}
date()

```


